---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### Loading and preprocessing the data

```{r} 
unzip("activity.zip")
initialData <- read.csv("activity.csv", colClasses=c("numeric", "Date", "numeric"))
```

Prilimnary EDA

```{r} 
head(initialData)
str(initialData)
```

### What is mean total number of steps taken per day?

Removing NAs

```{r}
data <- initialData[!(is.na(initialData$steps)), ]
```

Preparing aggreagted level data for the calculation

```{r}
totalStepsDay <- aggregate(steps ~ date, data, sum)
head(totalStepsDay)
```

Creating a histogram to indicate the frequency of total steps taken each day.


```{r}
paletteBlue <- colorRampPalette(c("red", "green", "blue"))
hist(totalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     main="Histogram of the Total Number of Steps Taken per Day",
     col=paletteBlue(22), family="serif")
```

The mean and median values of the total number of steps taken per day.

```{r, message=FALSE}
library(dplyr)
totalStepsSummary <- summarise(totalStepsDay, meanOfTotalSteps=mean(totalStepsDay$steps),
                               medianOfTotalSteps=median(totalStepsDay$steps))
print(totalStepsSummary)
```

The mean value calculated is **10766.19**, and the median value **10765**.

### Question 3: What is the average daily activity pattern?


Preparing aggreagted level data to split the data into groups for each interval, and then averaged with the mean function.

```{r}
meanStepsInterval <- aggregate(steps ~ interval, data, mean)
head(meanStepsInterval)
```

Time series plot (i.e. type = "l) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
plot(x=meanStepsInterval$interval, y=meanStepsInterval$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval",
     ylab="Number of Steps", xlab="Intervals (in 5 mins)",
     col="darkblue", lwd=1.5, family="serif")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
meanStepsInterval[grep(max(meanStepsInterval$steps), meanStepsInterval$steps), ]
```

The interval with the maximum number of steps is interval **835**.

### Question 4: Imputing missing values

As the data can be confirmed to contain some NA values as some of the observations:

```{r}
anyNA(initialData)
```

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NANAs)

```{r}
data.frame(steps=sum(is.na(initialData$steps)), 
           interval=sum(is.na(initialData$interval)), 
           date=sum(is.na(initialData$date)))
```

It can be seen that all 2304 NA values are contained within the steps variable.

Therefore an imputing strategy to replace all of these missing values with the mean value for the same interval, averaged across all days.

```{r}
imputedData <- initialData
for(x in 1:17568) {
    if(is.na(imputedData[x, 1])==TRUE) {
        imputedData[x, 1] <- meanStepsInterval[meanStepsInterval$interval %in% imputedData[x, 3], 2]
    }
}
head(imputedData)
```

Preparing aggregated data for the histogram

```{r}
imputedTotalStepsDay <- aggregate(steps ~ date, imputedData, sum)
head(imputedTotalStepsDay)
```

Creating the histogram:

```{r}
paletteRed <- colorRampPalette(c("red", "green", "blue"))
hist(imputedTotalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     main="Histogram of Total Number of Steps Taken per Day (With Imputed Values)",
     col=paletteRed(22), family="serif")
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
imputedStepsSummary <- summarise(imputedTotalStepsDay, 
                                 meanOfTotalSteps=mean(imputedTotalStepsDay$steps), 
                                 medianOfTotalSteps=median(imputedTotalStepsDay$steps))  
print(imputedStepsSummary)
```

This a similar calculation to the one completed at the start of the assignment, (without the imputed values,) and thus they can be compared.

```{r}
rbind(totalStepsSummary, imputedStepsSummary)
```

The values of the two data sets are very similar, if not exactly the same, due to the use of averaging functions when imputing the NA measurements. The mean values are the same, at **10766.19** steps, while the median value is slightly larger for the imputed data set, at **10766.19** steps, rather than **10765** steps.

If histograms of the two data sets (imputed and non-imputed) are compared:

```{r}
par(mfrow = c(1, 2))

hist(totalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     col=paletteBlue(22), family="serif", ylim=c(0, 20), main=NULL)
hist(imputedTotalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     col=paletteRed(22), family="serif", ylim=c(0, 20), main=NULL)
mtext("Histograms of Total Number of Steps Taken per Day, Without/With Imputed Values",
      adj=0.95, family="serif", font=2)
```

It can be seen that the frequency of values increases in the second histogram, which is expected, due to the imputed values. 


### Question 5: Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

Used the weekdays function to automatically calculate the day of the week each day resided upon,then the for loop which would assign the factor value "weekend" to all rows it read as having the values "Saturday" or "Sunday", and assign "weekday" to the others.

```{r}
daysData <- imputedData
daysData$days <- weekdays(daysData$date)
daysData$weekday <- as.character(rep(0, times=17568))
for(x in 1:17568) {
    if(daysData[x, 4] %in% c("Saturday", "Sunday")) {
        daysData[x, 5] <- "weekend"
    } else {
        daysData[x, 5] <- "weekday"
    }
}
daysData$weekday <- factor(daysData$weekday)
head(daysData)
```

The data has to be split into two groups of weekday/weekend data, using the newly created variable.

```{r}
weekdayData <- daysData[daysData$weekday=="weekday", ]
weekendData <- daysData[daysData$weekday=="weekend", ]
```

The average number of steps per interval is calculated, much like it has been done in previous questions.

```{r}
weekdayMean <- aggregate(steps ~ interval, weekdayData, mean)
weekendMean <- aggregate(steps ~ interval, weekendData, mean)
```

The panel plot is created. The x axis indicates each 5 minute interval, and the y axis shows the average number of steps taken. The two plots are divided into weekday, and weekend data.

```{r}
par(mfrow=c(2, 1), mar=c(4, 4.1, 3, 2.1))
plot(weekdayMean$interval, weekdayMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekdays",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps", family="serif",
     col="red", lwd=1.5, ylim=c(0, 230))
plot(weekendMean$interval, weekendMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekends",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps", family="serif",
     col="green", lwd=1.5, ylim=c(0, 230))
```
